--- 
title: "baby names" 
author: "nikita jayan" 
date: "3/12/2019" 
output: 
  word_document: default
  pdf_document: default
  html_document: default
--- 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#library(dplyr) # don't need to call this- dplyr is core tidyverse
library(tidyverse) 
``` 
# Make a title here please and edit everything at will In this analysis, I looked at ...
```{r}
# library(tidyverse) 
# install.packages('babynames') 
library(babynames) 
# install.packages('kableExtra') 
library (kableExtra)
``` 
## What does this code do? (Here's the code we ran successfully. If you like, come up with alternative ways of creating an appropriate dataset)

```{r} 
str(babynames) 
maleNames <- babynames %>% 
filter(sex == "M") %>% 
    rename (nBoy = n, propBoy = prop) 
femaleNames <- babynames %>% 
    filter(sex == "F") %>% rename (nGirl = n, propGirl = prop) 
androNames <- maleNames %>% 
inner_join(., femaleNames, by = c("name", "year")) 
``` 
## Can you fix this, or come up with something else? Here's the code that we started to write, but didn't finish. We were trying to find "the most androgynous names." You can continue to explore this, or find something new. Be clear, and finish with a brief table. Round numbers to just a few decimals, please. 
```{r} 
androNames %>% mutate (total = nGirl + nBoy) %>% 
mutate (sexRatio1 = nBoy*nGirl) %>% mutate (sexRatio2 = propBoy*propGirl) %>% # arrange(desc(total)) 
arrange(desc(sexRatio2)) %>% # you might mutate differently, and/or select certain cases 
head(10) %>% kable () %>% kable_styling() 
```
## What does it mean? 
The results from this table essentially state that Willie was the most androgynous name during the years 1900 and 1902-1910. 
Write a line or two describing this. Knit into a PDF, and be ready to upload on Wednesday. 
