
# Examining local COVID data in R

In Winter 2020, as the COVID-19 pandemic began to take hold, I began chronicling the course of the virus. What follows is an archive, followed by a consideration of some possible new directions. Note that the data reading section in the first few paragraphs is very slow.

## Tracking the Novel Coronavirus (from Feb 2020)

Here, I want to consider a timely (but challenging) dataset.

The Novel Coronavirus is an emerging health crisis, particularly in Wuhan (a Chinese city larger than New York City) and the surrounding province of Hubei. **It is not yet a threat in the United States - there have, at this writing (02/11/20) been zero cases in Florida.** Still, tracking the spread of the virus - the unfolding number of people infected and recovered, as well as the number of deaths - is a fascinating exercise.

This is an educational script for students learning R with the Tidyverse. It reads data provided by the Johns Hopkins Center for Systems Science and Engineering (JHU/CSSE).  

It was modified February 3 because of new GoogleSheet link and altered variable names, on Feb 5 because of a new URL for the data and additional changes in the variable name for date, and Feb 7 to (a) remove need for OAuth and (b) separate Wuhan from other China. On Feb 9, additional data cleaning was performed and interactive plots were added. On February 11, the code was rewritten to read files from a Github repo rather than Google Sheets.

This does not use an API or require authorization from Github.

```{r coronasetup, message = FALSE}
library(tidyverse)
library(magrittr)
library(lubridate)
library(plotly)
library(htmlwidgets)
library(httr)
library(purrr)
# get list of files
filelist <- GET("https://api.github.com/repos/CSSEGISandData/2019-nCoV/git/trees/master?recursive=1") %>% 
  content() %>% 
# there is probably a more efficient way to reduce this
# list to a set of filenames
  flatten() %>% 
  map ("path") %>%
  flatten() %>%
  tibble() %>% 
  rename(filename = 1) %>% 
  filter(str_detect(filename,".csv") &
           str_detect(filename,"daily")) 
nsheets <- nrow(filelist)
rawGitFiles <- "https://raw.githubusercontent.com/CSSEGISandData/2019-nCoV/master/"
```

### Reading the data

The Novel Coronavirus data consists of a series of csv files in a Github repository. This combines them into a single sheet in R. 

```{r readcoronadata, message = FALSE}
# variables to retain or create
numvars <- c("Confirmed", "Deaths", "Recovered")
varlist <- c("Province/State", "Country/Region",
             "Last Update", numvars) 
# one cool trick to initialize a tibble
coronaData <- varlist %>%
     map_dfr( ~tibble(!!.x := logical() ) )
# add data from files to tibble
for (i in 1:nsheets) {
  j <- read_csv(paste0(rawGitFiles,filelist$filename[i]))
# if a variable doesn't exist in sheet, add it
  j[setdiff(varlist,names(j))] <- NA
# datetime is formatted inconsistently
# across files, this must be done before merging  
  j %<>% mutate(`Last Update` = 
           parse_date_time(`Last Update`,
                           c('mdy hp','mdy HM',
                             'mdy HMS','ymd HMS'))) %>% 
  select(varlist) 
  coronaData <- rbind(coronaData, j) 
}
head(coronaData)
str(coronaData)
```

### Cleaning (wrangling, munging) the data (Feb 2020)

Cleaning the data includes not just finding "errors," but adapting it for our own use. It's generally time consuming, as was the case here. The following letters refer to sections of the code below.

* a - fix a few missing values outside of China for province and country
* b - the earliest cases, all in China, did not include country
* c - because province/state is included inconsistently, an unambiguous place variable is created
* d - reportdate is standardized (above) and renamed
* e - in some cases, multiple reports are issued for each day. only the last of these is used for each place.
* f - for dates where no data was supplied, the most recent (previous) data are used
* g - values of NA for Deaths, Confirmed, and Recovered cases are replaced by zero.
* h - Prior to Feb 1, 2020 reporting for US included only state, since then, city and state. This drops the (duplicated) province/state-only values beginning Feb 1. 

```{r cleaningcorona}
# the bidirectional pipe edits and replaces a dataset. use with caution.
coronaData <- coronaData %>% 
# a 
  mutate (`Province/State` = case_when(
    (is.na(`Province/State`) & 
       (`Country/Region` == "Australia")) ~ "New South Wales",
    (is.na(`Province/State`) & 
       (`Country/Region` == "Germany")) ~ "Bavaria", 
    TRUE ~ `Province/State`)) %>% 
  mutate (`Country/Region` = case_when(
    `Province/State` == "Hong Kong" ~ "Hong Kong",
    `Province/State` == "Taiwan" ~ "Taiwan",
    `Province/State` == "Washington" ~ "US",
# b
    is.na (`Country/Region`) ~ "Mainland China",
    TRUE ~ `Country/Region`)) %>% 
# c 
  mutate(place = ifelse(is.na(`Province/State`),
                        `Country/Region`,
                        paste0(`Province/State`,", ",
                               `Country/Region`))) %>% 
  mutate(reportDate = 
           date(`Last Update`)) %>% 
  group_by(place,reportDate) %>% 
# e
  slice(which.max(`Last Update`)) %>% 
  ungroup() %>%
# fill in missing dates for each place for time series
# f
  group_by(place) %>% 
  complete(reportDate = seq.Date(min(reportDate),
                                 today(),
                                 by="day")) %>% 
  fill(c(Confirmed,Deaths,Recovered,
         `Country/Region`,`Province/State`)) %>% 
# g
  ungroup() %>% 
  mutate_if(is.numeric, ~replace_na(., 0)) %>% 
# h
  mutate(dropcase = ((!str_detect(`Province/State`,",")) & 
                       (reportDate  > "2020-01-31") &
                       (`Country/Region` == "Canada" | `Country/Region` == "US"))) %>% 
# dplyr called explicitly here because plotly has taken over 'filter'
  dplyr::filter (!dropcase) %>% 
  select(-c(`Last Update`,`Province/State`,`Country/Region`,dropcase)) 
head(coronaData)
```

### 11 months later: The code still runs! 

The above code runs without apparent error, and leads to a dataset of `r nrow(coronaData)` rows by `r ncol(CoronaData)` columns. Here's a quick peek (Note that I use the group_by and summarize functions to collapse the file to one line per date):

```{r}
coronaData %>% group_by(reportDate) %>% 
  summarize(Confirmed = sum(Confirmed)) %>% 
  head() 
```

### Shall we graph it?

So far, so good. Let's plot the whole range. What do we see? This will generate a quick graph in ggPlot which shows the global incidence of confirmed cases. 

```{r}
coronaData %>% 
  group_by(reportDate) %>% 
  summarize(Confirmed = sum(Confirmed)) %>% 
  ggplot(aes(x=reportDate)) +
  geom_line(aes(y=Confirmed)) 
```
### Too bad
  
According to the above graph, there have been no additional COVID cases since mid-March or so. Too bad that the data aren't right here - for us and the world.
  
> **Exercise 8\_1** *Working with classmates, find the exact date when the file stopped updating. Use whatever method you like for this - kludgy or clever - but you should describe your results in explicit language or code so that anyone else would get the same results.*

### What now? 

I want to make sure that the rest of my code still works, so I'll look more closely at the valid data. I'll generate a new variable ('firstbaddate,' which sounds like a corny romcom), and filter by it. 

When the code was written a year ago, the COVID outbreak was largely contained to the Hubei province (which includes the city of Wuhan). So I tried breaking down the data into three locations, breaking down China into Hubei, other China, and the rest of the world.

And I generated summaries for three measures - Confirmed, Deaths, and Recovered cases

```{r threelocationscorona}

firstbaddate <- '2020-03-22' # This line is added in 2021
coronaDataSimple <- coronaData %>% 
  filter(reportDate < firstbaddate) %>%  # added in 2021
  mutate(country = case_when(
    str_detect(place,"China") ~ "China",
    TRUE ~ "Other countries")) %>% 
  mutate(location = case_when(
    place == "Hubei, Mainland China" ~ "Hubei (Wuhan)",
    country == "China" ~ "Other China",
# what happens when this line is not commented out?
# why is it written this way?
# str_detect(place, "ruise") ~ "Cruise Ship", 
    TRUE ~ "Elsewhere")) %>% 
  group_by(location,reportDate) %>% 
  summarize(Confirmed = sum(Confirmed),
            Deaths = sum(Deaths),
            Recovered = sum(Recovered)) %>% 
  ungroup()
```

### An initial plot: Data through February 11

The first plot is simple, including data for only deaths. A caption is added to show the source of the data.
Here's what the data looked like last February 11:

```{r simpleplot}
myCaption <- " Data courtesy JHU/CSSE http://bit.ly/ncvdata"
oldData <- coronaDataSimple %>% 
  filter(reportDate < '2020-02-12')
coronaPlot0 <- oldData %>% 
#  filter(reportDate < '2020-02-12')
  ggplot(aes(x=reportDate)) +
  geom_line(aes(y=Deaths, color = location)) +
  labs(caption = myCaption)
coronaPlot0
```

### five weeks later

About five weeks later our data would stop updating. But the world had already changed: Here's the same graph, through March 21, 2020: 

```{r simpleplot2 }
coronaPlot0 <- coronaDataSimple %>% 
  ggplot(aes(x=reportDate)) +
  geom_line(aes(y=Deaths, color = location)) +
  labs(caption = myCaption)
coronaPlot0
```

### Adding recovered cases (Feb 2020)

Here, recovered cases and deaths are included (as these are roughly on the same scale). Additional changes are self-evident.

```{r deathsrecovered}
mySubtitle <- paste0(
         "Recovered cases (solid line) and deaths (dotted) by region through ",
         firstbaddate)
 #        (month(today())), "/",
 #         (day(today())),"/",
 #         (year(today())),".")
myCaption <- " Data courtesy JHU/CSSE http://bit.ly/ncvdata"
coronaPlot1 <- coronaDataSimple %>% 
  ggplot(aes(x=reportDate)) +
  geom_line(aes(y=Recovered, 
                color = location), 
            linetype = "solid") + 
  geom_line(aes(y=Deaths, 
                color = location), 
            linetype = "dotted") +
  theme(axis.title.y = 
        element_text(angle = 90,
                     vjust = 1,size = 14),
        legend.position = (c(.2,.8))) +
  labs(title = "Novel coronavirus",
       subtitle = mySubtitle,
       y = "Cases", 
       caption = myCaption)
coronaPlot1
```

### Make the graph interactive

Plotly is an open-source, javascript based library that produces interactive graphs. The syntax that Plotly requires is (a little) different from ggplot, so, for example, the subtitle and caption are folded in to the title here, and the legend is moved a little further over. 

```{r confirmed}
p <- ggplotly(coronaPlot1) %>% 
  # make interactive
  layout(legend = list(x=.1,y=.9),
         title = list(text = paste0('Novel coronavirus',
                                    '<br>',
                                    '<sup>',
                                    mySubtitle,
                                    myCaption,
                                    '</sup>')))
# saveWidget(p, file="coronaDeathsRecovered.html")
p
```

### Plotting confirmed cases

In this last figure, data for confirmed cases are shown (only the interactive version is included here): 

```{r}
mySubtitle <- paste0(
         "Confirmed cases by region through ",
         (month(today())), "/",
         (day(today())),"/",
         (year(today())),".")
coronaPlot2 <- coronaDataSimple %>% 
  ggplot(aes(x=reportDate)) +
  geom_line(aes(y=Confirmed, 
                color = location), 
            linetype = "solid") + 
  theme(axis.title.y = 
        element_text(angle = 90,
                     vjust = 1,size = 14),
        legend.position = (c(.2,.8))) +
  labs(title = "Novel coronavirus",
       subtitle = mySubtitle,
       y = "Cases", 
       caption = myCaption)

p <- ggplotly(coronaPlot2) %>% 
  # make interactive
  layout(legend = list(x=.1,y=.9),
         title = list(text = paste0('Novel coronavirus',
                                    '<br>',
                                    '<sup>',
                                    mySubtitle,
                                    myCaption,
                                    '</sup>')))
# saveWidget(p, file="coronaConfirmed.html")
p
```

### Some questions

1. Consider the data and try to run the code yourself.
  + What problems did you encounter? 
  + What parts need to be annotated more?

2. Can you reverse-engineer my code? Where is it confusing? (remember the 15 minute rule).

3. What is the relationship between 'confirmed cases' and 'deaths'? Which appears to be increasing more quickly?  Is this cause for optimism?

4. Can you improve on these plots? 

5. Some more challenging questions.
  + What is (roughly) the shape of the function for each of the three variables, and for China/Other? 
  + What values would you expect for, say, ten days from now?

### Additional notes

If you are interested in looking at additional epidemiological datasets and how they might be looked at in R, consider this source by Tomás J. Aragón (https://bookdown.org/medepi/phds/). For Plotly in R, check out https://plotly-r.com/ 

